---
title: "Class 11: Structural Bioinformatics (Part 1)"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Revisit the PDB

Q1. Open RStudio and begin a new class11 project within your GitHub tacked directory/folder from last day. Make sure “Create a git repository” option is NOT ticked. This is because we want to use the same git repository as we used last day and not start a new one - if you are not sure what this means ask Barry now!

Download a CSV file from http://www.rcsb.org/stats/summary on 2019-05-08

```{r}
db <- read.csv("Data Export Summary.csv", row.names = 1)
head(db)
```

```{r}
## Total number of structures in the PDB Database
sum(db$Total)

## Percent of each experiment method
rowSums(db[,1:4])

(rowSums(db[,1:4])/sum(db$Total)) * 100
```

Clean version of the percentage for each experimental method (one liner)
```{r}
db$Total/sum(db$Total)*100
```

What percent are protein?
```{r}
(sum(db$Proteins)/sum(db$Total))*100
```



## Section 3. Using Bio3D

Let's read the PDB file we downloaded into R to see how things live in this place

```{r}
library(bio3d)
```


```{r}
pdb <-  read.pdb("1hsg.pdb")
pdb
pdb$seqres
```

```{r}
aa321(pdb$seqres)
```


```{r}
attributes(pdb)
```


```{r}
ca.inds <- atom.select(pdb, "calpha")
```

Let's select residue 10
```{r}
inds <- atom.select(pdb, resno=10)
```

```{r}
pdb$atom[inds$atom,]
```


## Make a protein only and ligand only PDB file

Here we need to select our protein with 'atom.select()' then use 'write.pdb()'

```{r}
prot <- atom.select(pdb, "protein", value = TRUE)
prot

```

```{r}
write.pdb(prot, file="1hsg_protein.pdb")
```


```{r}
lig <- atom.select(pdb, "ligand", value = TRUE)
lig
```

```{r}
write.pdb(lig, file = "1hsg_ligand.pdb")
```


## 3D Structure Viewing in R
```{r}
# The 'devtools' package allows us to install development versions
# install.packages("devtools")

# Install the bio3d.view package from bitbucket
# devtools::install_bitbucket("Grantlab/bio3d-view")

library("bio3d.view")
#install.packages("rgl")
view(pdb, "overview", col="sse")
```

```{r}
library(bio3d.view)
# Structural displacements captured by PC1
#view(pc1)
# The rglwidget() function from the rgl # package will show output in your Rmd 
# notebook and rendered html_output documents
#library(rgl) 
#rglwidget(pc1)
```



## Section 6: working with multiple pdb files

```{r}
# Download some example PDB files
ids <- c("1TND_B","1AGR_A","1TAG_A","1GG2_A","1KJY_A","4G5Q_A") 

files <- get.pdb(ids, split = TRUE)
files
```


```{r}
# Extract and align the chains we are interested in
pdbs <- pdbaln(files, fit = TRUE)
# Print to screen a summary of the 'pdbs' object
view(pdbs)
```


```{r}
## Align the first 5 rows and 8 columns
pdbs$ali[1:5, 1:8]

# Associated residues numbers
pdbs$resno[1:5, 1:8]
```



```{r}
# Calculate sequence identity
seqidentity(pdbs)
```


```{r}
# Calculate RMSD
rmsd(pdbs)
```



```{r}
# Calculate RMSD
rd <- rmsd(pdbs)

# Clustering
hc <- hclust(as.dist(rd))
grps <- cutree(hc, k=3)

# Plot results as dendrogram
hclustplot(hc, k=3)
```

In Section 7 we walk through a much more detailed and large-scale PCA analysis using all available structures for a given protein family. Here for example purposes we will perform PCA on the already loaded pdbs object with the function pca.xyz().

```{r}
# Perform PCA
pc.xray <- pca(pdbs)
# Plot our results summary (PCA score plot and scree-plot)
plot(pc.xray)
```


```{r}
# Visualize first principal component
pc1 <- mktrj(pc.xray, pc=1, file="pc_1.pdb")
```


