---
title: "Class 12: Structural Bioinformatics (Part 2)"
author: "Frank Lee"
date: "5/10/2019"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Obtain and prepare our structure for 'docking'

Here we download the HIV-Pr structure from the PDB database and extract out the protein and ligand portions to separate new PDB format files. 

# In here, we have a few non-protein/nucleic resid values: 127 water and 1 MK1
# We have 172 non-protein/nucleic atoms
```{r}
library(bio3d)

file <- get.pdb("1hsg")
pdb <-  read.pdb(file)
pdb
```

# In this one, we don't have any non-protein/nucleic resid values
# We also don't have any non-protein/nucleic atoms
```{r}
# Select protein only
prot <-  atom.select(pdb, "protein", value = TRUE)
prot
write.pdb(prot, file="1hsg_protein.pdb")
```

# For ligand, we have 1 non-protein/nucleic resid value: MK1
# We also have 45 non-protein/nucleic atoms
```{r}
# Select ligand only
lig <-  atom.select(pdb, "ligand", value = TRUE)
lig
write.pdb(lig, file="1hsg_ligand.pdb")
```

## Add Hydrogens and charges in ADT

We opened our protein-only PDB file in AutoDocTools and added hydrogens and atom-types needed for docking calculations. 

## Run docking

We will use AutoDoc Vina here at the UNIX command line!

The command I used in Terminal was: 
> ~/Downloads/autodock_vina_1_1_2_mac/bin/vina --config config.txt --log log.txt


## Process results back in R

```{r}
# Set multi to true because there are multiple answers
res <- read.pdb("all.pdbqt", multi = TRUE)
res

# Now let's write res as a new pdb as a file named 'results.pdb'
write.pdb(res, file="results.pdb")
```

This tells us that there are 14 models. 



Compare our docking poses to the MERK drug structure. 

```{r}
ori <- read.pdb("ligand.pdbqt")
rmsd(ori, res)
```

The smaller the value in rmsd (root mean square deviation), the smaller the distance. 


## Normal Mode Analysis example
Normal Mode Analysis (NMA) is a bioinformatics method that can predict the major motions of biomolecules.

```{r}
pdb <- read.pdb("1hel")
modes <- nma( pdb )
m7 <- mktrj(modes, mode=7, file="mode_7.pdb")
```

```{r}
plot(modes)
```


Then you can open the resulting mode_7.pdb file in VMD - Use "TUBE" representation and hit the play button...
Or use the bio3d.view view() function

```{r}
library("bio3d.view")
view(m7, col=vec2color(rmsf(m7)))
```

