---
title: "Class 4: Bioinformatics Data Analysis with R"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Background:
R is powerful data programming language that you can use to explore and understand data in an open-ended, highly interactive, iterative way. Learning R will give you the freedom to experiment and problem solve during data analysis — exactly what we need as bioinformaticians.

Simple Calculations
```{r}
5+3
5-3
5*3
5/3

```

Object names cannot start with a digit and cannot contain certain other characters such as a comma or a space.

```{r}
x <- 3*4
x
```

```{r}
seq(1,10)
```

```{r}
seq(1,10, by=2)
```

```{r}
example(seq)
```

Help function
```{r}
help(log)
```

Vectors: a container of contiguous data. Values are not stored as scalar; rather, as a vector of length 1. 
```{r}
length(3.1)
```

```{r}
x <- c(56, 95.3, 0.4)
y <- c(3.2, 1.1, 0.2)

x
y

x+y
x-y
x/y
```


```{r}
sqrt(x)
round(sqrt(x),3)
log(x)/2 + 1

x[2]
x[3] <- 0.5
x
```

```{r}
x <- 1:50
plot(x, sin(x), typ="l", col="red")

```

```{r}
plot(x, sin(x), typ="l", col="blue", lwd=3, xlab="x = 1:50")
```

