---
title: "Class 16: Introduction to UNIX"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(bio3d)
```


```{r}
tsv <- read.csv("mm-second.x.zebrafish.tsv", sep = "")
head(tsv)
```

Make a histogram of the $bitscore values. You may want to set the optional breaks to be a larger number (e.g. breaks=30).

```{r}
colnames(tsv)
colnames(tsv) <- c("qseqid", "sseqid", "pident", "length", "mismatch", "gapopen", "qstart", "qend", "sstart", "send", "evalue", "bitscore")

hist(tsv$bitscore)
hist(tsv$bitscore, breaks = 30)
```
The answer is that bitscores are only somewhat related to pident; they take into account not only the percent identity but the length of the alignment. You can get a napkin sketch estimate of this by doing the following:

```{r}
plot(tsv$pident  * (tsv$qend - tsv$qstart), tsv$bitscore)
```

Or using ggplot (note you will have to install the ggplot2 package first if you are working on JetStream)

```{r}
library("ggplot2")
```

```{r}
ggplot(tsv, aes(pident, bitscore)) + geom_point(alpha=0.1) 

```

```{r}
ggplot(tsv, aes((tsv$pident * (tsv$qend - tsv$qstart)), bitscore)) + geom_point(alpha=0.1) + geom_smooth(aes((tsv$pident * (tsv$qend - tsv$qstart)), bitscore))

```

## Part 3

```{r}
shmlast_out <- read.csv("mouse.1.rna.fna.gz.x.cow.faa.crbl.csv")
head(shmlast_out)
```


```{r}
View(shmlast_out)
dim(shmlast_out)
```

Let’s do some data visualization to get a handle on what our blast output looked like: first, let’s look at the E_scaled column.

```{r}
hist(shmlast_out$E_scaled)
```

We can take a look at some more stats – let’s look at the bitscore column:
```{r}
hist(shmlast_out$bitscore)
```

We can also look at the length of the queries, which are the mouse sequences in this case.
```{r}
hist(shmlast_out$q_len)
plot(shmlast_out$q_len, shmlast_out$bitscore)
```

```{r}
shmlast_sub = shmlast_out[sample(nrow(shmlast_out), 10),]
View(shmlast_sub)
```

```{r}
tmp <- subset(shmlast_out, q_len >= 8000 & q_len <= 11000 & bitscore <=2000)
functions <- tmp[, c("q_name", "s_name")]
```

